### 2019-01-21 Datelife_paper1.Rmd
finished description and updating refs

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---
```{r eval = TRUE, echo = FALSE}
fig_workflow_main <- "Fig. 1"
fig_runtime_main <- "Fig. 2"  
# make_datelife_query_plus_get_datelife_result
```

@@

Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across
various repositories (otol tree store, dryad, treebase, journals supplementary data)
usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
<!--General issues with dating techniques:--> choosing correct fossils for calibration r
equires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.
<!--INTRODUCTION: 5th paragraph-->

DateLife palliates this by only using information available from already published
studies, which are ideally constructed using robust information, such as sequence
data and thoughtfully curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->
DateLife can summarize this information in several formats that can be easily inspected
by users.
This allows rapidly obtaining a time frame of lineage divergence for a wide number
of taxa.
DateLife can also generate chronograms for taxa with little available information,
by using the available data as calibration points.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package <link to documentation>, a web interface
(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
In here we present the first release of DateLife.
It contains an improved databse of chronograms, more methods to summarize trees,
and new functions to visualize data.
It also allows comparison of summary methods.

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>
```{r, echo = FALSE}
fig_workflow_main_foot <- "Figure 1. Summarized DateLife workflow. Only main actions are shown."
```
<!--(BOLD and OToL species names, are they homogeneous/standardized?-->

DateLife is a service for searching and processing information on ages of any number
of taxa <!--or group of taxa-->of interest, across chronograms available in public
data repositories coming from published peer reviewed studies. It can also generate
new taxon age information<!--by using expert available phylogenies and scaling them
to absolute time using information on ages obtained from published chronograms,--> by l
inking several external services and tested algorithms.
It takes advantage of the rotl, ape, and geiger packages to gather, process, and
present information.

It only requires a set of taxon names as input, in the form of a comma separated
listing or vector, or of a phylogeny with taxon names on the tips.
Taxon names can correspond to binomial species names or clades
<!--common names (will we implement it?)-->.
When taxon names are clades, DateLife pulls all accepted species names within the
clade (up to OToL's limit of _____ species) from OToL's reference taxonomy using
a service of rphylotastic R package<!--citation-->.
Names belonging to subspecies or any other infraspecific category are treated as
species.
DateLife can process input names with the taxon name resolution service (TNRS),
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects misspelled names or typos, and standardizes variation in spelling
and synonyms [@Boyle2013], increasing the probability to correctly find the queried
taxa in the chronogram database. DateLife uses TNRS to compare names against OToL's
reference taxonomy using a service from the R package rotl [@Michonneau2016].
<!--we'll implement the option to use other reference taxonomies in the future -->

DateLife's main function searches taxon names across the chronogram database specified
by the user. At the moment, it queries chronograms from OToL [@Hinchliff2015]
<!-- it cannot query chronograms from TreeBASE -there is an earlier reference from
1994 but it does not exist in its journal database: Sanderson MJ, Donoghue MJ, Piel
WH, Eriksson T: TreeBASE: a prototype database of phylogenetic analyses and an interactive
tool for browsing the phylogeny of life. Am J Bot. 1994, 81 (6): 183--> repository.
DateLife identifies chronograms having at least two taxon names, and subset them
to contain only the taxa of interest. It then stores taxon age information from
each chronogram individually as a patristic matrix, named with the citation of the
original study. This format allows a rapid summary in a number of different ways,
including:
1) citations of the original studies containing the subset chronograms,
2) a list of mrca ages of subset chronograms,
3) a list of complete subset chronograms in newick or phylo format,
4) a table containing all information retrieved in html or R's data frame format, or
5) a single chronogram summarized from subset chronograms using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths. <!--(how to better explain the latter?? Also, I just discovered the median gives non ultrametric trees...).-->

DateLife also stores information on input taxon presence/absence across subset chronograms.
Users can choose to add ages of missing taxa to subset chronograms in different
ways, depending on the amount of knowledge they want to input or how much they want
to be involved in the steps of the addition process.
If users have no access to biological information (i.e., a character, DNA or protein
matrix), missing taxa can be added to any chronogram simply at random, or by following
taxonomic or phylogenetic knowledge from expert sources. There are a wide number
of open reference taxonomies available, such as the Catalogue of Life [@Roskov2017]
or the NCBI taxonomy database [@Federhen2012]. Expert phylogenies (with or without
branch lengths) to be used as topological constraint (backbone) can also be obtained
from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002]
and Dryad (<https://www.datadryad.org//>).
At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as
expert knowledge to automatically add missing taxa to chronograms. Alternatively,
users can input a reference taxonomy or topological constraint of their choosing
or making. <!-- at some point we will implement the possibility for users to choose
different reference taxonomies automatically-->
If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest,
DateLife can construct a sequence data matrix from DNA markers available from the
Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve
polytomies. It will follow OToL's synthetic tree as backbone.
<!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed?? To estimate node ages chronosMPL uses mean path length method from Britton et al. 2002-->
<!--It can rapidly construct trees with branch lengths from
a set of lineages with sequence data from the Barcode Of Life Database (BOLD, <http://www.boldsystems.org/>), using the synthetic OToL (<https://tree.opentreeoflife.org>) as backbone, and to obtain divergence dates.
It allows direct comparison of dates obtained with different markers
available in BOLD (in plants and fungi in particular).-->
To use information from a topological constraint, DateLife calls the congruification
method described in [@Eastman2013] to find shared nodes between trees (congruent
nodes). It then fixes their ages, and add ages to remaining nodes with a dating
method that can be specified by the user. If users have access to biological data,
they can input a tree with branch lengths proportional to relative substitution
rates as topological constraint. In this case, age data from congruent nodes will
be used as calibration points.
Age data from several chronograms can be combined and congruified to be used as
calibration points in a single analysis.

Several dating methods are implemented in DateLife.
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
DateLife implements BLADJ from the development R version of phylocom's R package
[@Webb2008], phylocomr (<https://github.com/ropensci/phylocomr>). It can only be
used when there is a topological constraint with no branch lengths.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is
also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012]. It is called through R.
MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
random, following a reference taxonomy or a topological constraint. It draws ages
from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
DateLife calls MrBayes trough an R function.

DateLife can also correct negative branch lengths in several ways.

@@

Supplementary material, including code files and
online-only appendices, can be found in the GitHub
repository <!-- Dryad also? since it is all in github already-->

@@

@@

```{r pressure, out.width="100%", include=FALSE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
knitr::include_graphics("fig1.pdf", dpi = 360)
```

>>>>

---
output: pdf_document
fontsize: 12pts
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
# indent: true # this works
# header-includes:
#    - \setlength\parindent{24pt}  # this works too, but I'mg onna indent manually with 4 spaces
# nope, 4 spaces indents are not recognized, they do something else
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
---
```{r eval = TRUE, echo = FALSE}
fig_workflow_main <- "Fig. 1"
fig_workflow_all_functions <- "Vignette Fig. 1"
fig_runtime_main <- "Fig. 2"  
# make_datelife_query_plus_get_datelife_result
```

@@

Despite its great importance, analytical tools to summarize available information
on taxon ages for the scientific community are still lacking.
We identified several aspects that might have so far delayed the exploitation of
existing data. First, original chronograms available publicly are scattered across
various repositories (otol tree store, dryad, treebase, journals supplementary data)
usually with different formats too.
Second, lineage names due to taxonomic idiosincracy can be different among studies
and manual curation of that is usually necessary.
Third, data curation
Recent advances on this area (e.g., supersmart<!--and, which others?-->) aim to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of
data curation and knowledge on the group of interest is still necessary. For example,
<!--General issues with dating techniques:--> choosing correct fossils for calibration 
requires a lot of expertise and knowledge on the group. An incorrect use of fossils
can generate severe bias in dating results [@Sauquet2012c].
<!--Data curation is necessary at some point (is this always true?)-->
Hence, data curation is still an important part of any biological study. The research
community considers it as an important or even crucial step before data analysis.
Hence, automated processes for large data analysis are frequently received with skepticism.
<!--INTRODUCTION: 5th paragraph-->

DateLife palliates this by only using information available from already published
studies, which are ideally constructed using robust information, such as sequence
data and thoughtfully curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->
DateLife can summarize this information in several formats that can be easily inspected
by users.
This allows rapidly obtaining a time frame of lineage divergence for a wide number
of taxa.
DateLife can also generate chronograms for taxa with little available information,
by using the available data as calibration points.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system
[@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package <link to documentation>, a web interface
(<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.
In here we present the first release of DateLife.
It contains an improved database of chronograms, more methods to summarize trees,
and new functions to visualize data.
It also allows comparison of summary methods.

<!--DESCRIPTION of Datelife-->
<span style="font-variant:small-caps;">
Description
</span>
```{r, echo = FALSE}
fig_workflow_main_foot <- paste("Stylized DateLife workflow. Detailed steps are shown in",  fig_workflow_all_functions)
```
<!--(BOLD and OToL species names, are they homogeneous/standardized?-->
The basic `DateLife` workflow is shown in `r fig_workflow_main` and consists of:

1) A user providing as input at least two taxon names, either as tip labels on a tree, or as a simple comma separated character string. The tree can be in newick or phylo format, and can be with or without branch lengths.
2) `DateLife` then performs a search across its database of peer reviewed and curated chronograms; identifies and gets source trees with at least two matching input names; drops unmatching taxa from positively identified source trees; and finally transforms each source tree to a patristic matrix named by the citation of the original study. This format facilitates and greatly speeds up all further analyses and summary algorithms.
3) The user can obtain different types of summaries from the source data including: a) all source chronograms, b) mrca ages of source chronograms, c) citations of studies where source chronograms were originally published, d) a summary table with all of the above, e) a single summary tree of all source chronograms, and f) a report of succesful matches per input taxon name across source chronograms.
4) At this point, users can choose to use all or some source data as calibration points to date a tree of their own making or choosing. <!--, a taxonomic tree-->
5) Users can also simulate age and/or phylogenetic data of input taxa not found in the database. A variety of algorithms are available for this purpose.
6) Finally, users can easily view results graphically as well as construct their own graphs using inbuilt `DateLife` graphic generators.

<!--To perform the former tasks:-->
`DateLife`'s chronogram database is currently built from Open Tree of Life (OToL)'s [@Hinchliff2015] tree repository. Among currently existing repositories (e.g., TreeBAse, Dryad), OToL's metadata rich tree store  is the only one meeting the requirements for proper/accurate automatized handling of trees.
Input taxon names accepted by `DateLife` are binomial species names or clades. <!--common names (implemented yet?)--> Taxon searches are performed at the species level, so when input names correspond to higher clades, `DateLife` pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search.
Input names belonging to subspecies or any other infraspecific category are treated as
species.
DateLife also processes input names with the taxon name resolution service (TNRS),
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects potentially misspelled names and typos, and standardizes variation in spelling
and synonyms <!--to a taxonomic reference of users choosing, degfault to otol taxonomy -->[@Boyle2013], increasing the probability to correctly find the queried
taxa in `DateLife`'s chronogram database.

Source chronogram summary tree can be assembled using the Super Distance Matrix (SDM) supertree construction approach [@Criscuolo2006] or using the median of branch lengths and the hierarchical clustering method <!-- of ttol-->. <!--Also, I just discovered the median gives non ultrametric trees... but I thonk it is fixed now, check it!).--> 
Tree dating and simulation options are performed with various algorithms:
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated
nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. 
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012].
MrBayes [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
random, following a reference taxonomy or a topological constraint. It draws ages
from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
To apply calibrations to a tree, congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes).

To gather, process, and present information, `DateLife` builds up from functions available in several R packages including rotl [@Michonneau2016], ape[@Paradis2004], geiger[@Harmon2008], paleotree[@Bapst2012a], bold[@Chamberlain2018], phytools[@Revell2012], taxize[@Chamberlain2013; @Chamberlain2018], phyloch[@Heibl2008], phylocomr[@Ooms2018] and rphylotastic[@Omeara2019].

Details on each step are further developed in `DateLife`'s R package documentation `datelife workflow` vignette at (<https://LINK>).

@@

Supplementary material, including code files and
online-only appendices and vignettes, can be found in the GitHub
repository of the paper at (<https://github.com/LunaSare/datelife_paper1>) as well as in the package vignettes <!-- Dryad also? since it is all in github already-->
and are also available from the Dryad Digital Repository at LINK.

@@

```{r pressure, out.width="100%", include=TRUE, fig.align="center", fig.cap=fig_workflow_main_foot, echo=FALSE}
knitr::include_graphics("fig1.pdf", dpi = 360)
```

@@
